---
title: "Investigating the Association Between Maternal Folic Acid Supplementation During Pregnancy and the Risk of Low Birth Weight"
subtitle: "BMIN503/EPID600 Final Project"
author: "Michaela Olabisi"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers. Feel free to change the theme and other display settings, although this is not required. I added a new sentence

## Overview {#sec-overview}

Give a brief a description of your project and its goal(s), what data you are using to complete it, and what two faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

This project explores the relationship between maternal folic acid supplementation during pregnancy and the risk of delivering a low birth weight (LBW) infant, defined as a birth weight below 2,500 grams.^1^ Using data from the Pregnancy Risk Assessment Monitoring System (PRAMS), I aim to assess whether folic acid intake during pregnancy can help lower the likelihood of LBW. I consulted with Dr. John Holmes and Dr. Ellen Caniglia from the Department of Biostatistics, Epidemiology & Informatics. Dr. Holmes highlighted potential study limitations, such as selection bias, while Dr. Caniglia shared a relevant [paper](https://pubmed.ncbi.nlm.nih.gov/35561720/) that examines the link between supplementation and adverse pregnancy outcomes like preterm birth (PTB) and LBW. You can access the final project, including all code and results, on my GitHub: <https://github.com/msaolabisi/BMIN503_Final_Project.git>

## Introduction {#sec-introduction}

Describe the problem addressed, its significance, and some background to motivate the problem. This should extend what is in the @sec-overview.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

The problem I'm addressing in this project is the potential link between maternal folic acid supplementation during pregnancy and the risk of delivering LBW infant, defined as a birth weight under 2,500 grams. LBW is an important indicator of adverse pregnancy outcomes, including higher rates of neonatal morbidity, mortality, and long-term developmental issues like cardiovascular and metabolic disorders.^2^ These conditions put a lot of stress on healthcare systems, driving up costs and the need for more intensive care, including extended stays in the NICU (Neonatal Intensive Care Unit).^3^ It's important to identify maternal factors that can be changed to lower the risk of LBW, as this could help improve the health of both mothers and babies.

Folic acid is a B vitamin that’s often recommended during pregnancy to help prevent birth defects like spina bifida.^4^ It helps with DNA production, cell growth, and tissue development, which are all crucial for the rapid development of the fetus.^5^ Folic acid may also support the placenta, which provides the fetus with nutrients and oxygen, and a healthy placenta can help prevent LBW.^5^ By reducing these issues, folic acid might help lower the chances of having an LBW baby.

Despite these potential benefits, the direct relationship between folic acid and LBW isn’t fully established, and the research results are inconsistent. This is why it’s important to further investigate how folic acid supplementation during pregnancy might affect LBW.

This issue spans multiple fields. Epidemiology helps us understand the patterns and causes of LBW, while biostatistics is used to analyze large datasets like PRAMS. Public health looks at how factors like folic acid impact maternal and fetal health at the population level. Nutrition explains the role of folic acid in fetal development, and obstetrics and maternal health offer insights into how maternal health, prenatal care, and pregnancy complications influence birth weight outcomes.

When I spoke with Dr. John Holmes and Dr. Ellen Caniglia from the Department of Biostatistics, Epidemiology & Informatics, they helped me better understand the challenges of studying this issue. Dr. Holmes stressed the importance of addressing biases, like selection bias, when interpreting data in this case and Dr. Caniglia shared her research on supplementation and pregnancy outcomes, providing valuable context for studying the link between folic acid and LBW.

By drawing from these different fields of expertise, this project seeks to clarify the potential role of folic acid supplementation in reducing LBW.

## Methods {#sec-methods}

Describe the data used and general methodological approach used to address the problem described in the @sec-introduction. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.

For this secondary analysis, I'm using data from PRAMS, Phase 8, a surveillance system that collects state-specific data on maternal behaviors, experiences, and health outcomes before, during, and after pregnancy. The exposure for this study is folic acid supplementation (categorized as Yes/No) and the outcome is infant birth weight, which has been grouped into 250-gram intervals.

### **Exposure**

+-------------+------------------------+------------+-------------+
| PRE_VIT_RAW | Pre-preg -- folic acid | NYS1F      | .B=DK/BLANK |
|             |                        |            |             |
|             |                        |            | .S=SKIP     |
|             |                        |            |             |
|             |                        |            | 1=NO        |
|             |                        |            |             |
|             |                        |            | 2=YES       |
+-------------+------------------------+------------+-------------+

+-------------+-------------------+------------+-------------+
| VPP_VIT_RAW | PPV -- folic acid | NYS1F      | .B=DK/BLANK |
|             |                   |            |             |
|             |                   |            | .S=SKIP     |
|             |                   |            |             |
|             |                   |            | 1=NO        |
|             |                   |            |             |
|             |                   |            | 2=YES       |
+-------------+-------------------+------------+-------------+

### **Outcome**

+--------------+----------------------------------------------+-----------+----------------+
| GRAM_NAPHSIS | Birth weight grouped into 250-gram intervals | NP_GRAM   | .U=UNKNOWN     |
|              |                                              |           |                |
|              |                                              |           | 125=0227-0249  |
|              |                                              |           |                |
|              |                                              |           | 375=0250-0499  |
|              |                                              |           |                |
|              |                                              |           | 625=0500-0749  |
|              |                                              |           |                |
|              |                                              |           | 875=0750-0999  |
|              |                                              |           |                |
|              |                                              |           | 1125=1000-1249 |
|              |                                              |           |                |
|              |                                              |           | 1375=1250-1499 |
|              |                                              |           |                |
|              |                                              |           | 1625=1500-1749 |
|              |                                              |           |                |
|              |                                              |           | 1875=1750-1999 |
|              |                                              |           |                |
|              |                                              |           | 2125=2000-2249 |
|              |                                              |           |                |
|              |                                              |           | 2375=2250-2499 |
|              |                                              |           |                |
|              |                                              |           | 2625=2500-2749 |
|              |                                              |           |                |
|              |                                              |           | 2875=2750-2999 |
|              |                                              |           |                |
|              |                                              |           | 3125=3000-3249 |
|              |                                              |           |                |
|              |                                              |           | 3375=3250-3499 |
|              |                                              |           |                |
|              |                                              |           | 3625=3500-3749 |
|              |                                              |           |                |
|              |                                              |           | 3875=3750-3999 |
|              |                                              |           |                |
|              |                                              |           | 4125=4000-4249 |
|              |                                              |           |                |
|              |                                              |           | 4375=4250-4499 |
|              |                                              |           |                |
|              |                                              |           | 4625=4500-4749 |
|              |                                              |           |                |
|              |                                              |           | 4875=4750-4999 |
|              |                                              |           |                |
|              |                                              |           | 5125=5000-5249 |
|              |                                              |           |                |
|              |                                              |           | 5375=5250-5499 |
|              |                                              |           |                |
|              |                                              |           | 5625=5500-5749 |
|              |                                              |           |                |
|              |                                              |           | 5875=5750-5999 |
|              |                                              |           |                |
|              |                                              |           | 6125=6000-6249 |
|              |                                              |           |                |
|              |                                              |           | 6375=6250-6499 |
|              |                                              |           |                |
|              |                                              |           | 6625=6500-6749 |
|              |                                              |           |                |
|              |                                              |           | 6875=6750-6999 |
|              |                                              |           |                |
|              |                                              |           | 7125=7000-7249 |
|              |                                              |           |                |
|              |                                              |           | 7375=7250-7499 |
|              |                                              |           |                |
|              |                                              |           | 7625=7500-7749 |
|              |                                              |           |                |
|              |                                              |           | 7875=7750-7999 |
|              |                                              |           |                |
|              |                                              |           | 8125=8000-8165 |
+==============+==============================================+===========+================+
|              |                                              |           |                |
+--------------+----------------------------------------------+-----------+----------------+

### **Covariates**

### **Maternal Factors**

-   **Age** (categorized into age groups: .U=Unknown 17=17 or younger 45=45 or older) **MAT_AGE_PU**

-   **Parity** (categorized as: .U=UNKNOWN 0=0 1=1 2=2 3=3-5 4=6+) **PRE_LB_NAPHSIS**

-   **BMI (kg/m²)** (categorized using WHO classification): B=DK / BLANK, 1=UNDERWT (   \< 18.5), 2=NORMAL  (18.5-24.9), 3=OVERWT  (25.0-29.9), 4=OBESE   (30.0 +   ) **MOM_BMIG_QX_REV**

### **2. Socioeconomic Determinants**

-   **Education** (categorized as: .U=UNKNOWN 1=\<= 8TH GRADE 2=9-12 GRADE, NO DIPLOMA 3=HIGH SCHOOL GRAD / GED 4=SOME COLLEGE, NO DEG / ASSOCIATE DEG 5=BACHELORS / MASTERS / DOCTORATE / PROF) **MAT_ED**

-   **Place of residence** (STATE NAME) **STATE**

-   Income (Income -- 12 mnths bef, total income **INCOME8**

    .B=DK/BLANK

    1=\$ 0 TO \$16,000

    2=\$16,001 TO \$20,000

    3=\$20,001 TO \$24,000

    4=\$24,001 TO \$28,000

    5=\$28,001 TO \$32,000

    6=\$32,001 TO \$40,000

    7=\$40,001 TO \$48,000

    8=\$48,001 TO \$57,000

    9=\$57,001 TO \$60,000

    10=\$60,001 TO \$73,000

    11=\$73,001 TO \$85,000

    12=\$85,001 OR MORE

    13=\$ 0 TO \$10,000

    14=\$10,001 TO \$16,000

    15=\$85,001 TO \$99,999

    16=\$100,000 OR MORE

    17=\$85,001 TO \$99,000

    18=\$99,001 TO \$109,000

    19=\$109,001 OR MORE

    20=\$ 85,001 TO \$100,000

    21=\$100,001 TO \$120,000

    22=\$120,001 OR MORE

    101=\$ 0 TO \$ 20,000

    102=\$20,001 TO \$ 25,000

    103=\$ 25,001 TO \$ 30,000

    104=\$ 30,001 TO \$ 36,000

    105=\$ 36,001 TO \$ 40,000

    106=\$ 40,001 TO \$ 50,000

    107=\$ 50,001 TO \$ 60,000

    108=\$ 60,001 TO \$ 71,000

    109=\$ 71,001 TO \$ 75,000

    110=\$ 75,001 TO \$ 91,000

    111=\$ 91,001 TO \$107,000

    112=\$107,001 OR MORE

    201=\$ 0 TO \$ 18,000

    202=\$18,001 TO \$ 23,000

    203=\$23,001 TO \$ 28,000

    204=\$28,001 TO \$ 33,000

    205=\$33,001 TO \$ 37,000

    206=\$37,001 TO \$ 46,000

    207=\$46,001 TO \$ 55,000

    208=\$55,001 TO \$ 65,000

    209=\$65,001 TO \$ 69,000

    210=\$69,001 TO \$ 84,000

    211=\$84,001 TO \$ 98,000

    212=\$98,001 OR MORE) INCOME8F

### **3. Utilization of Antenatal Care (ANC)**

-   **PNC_VST** Number of prenatal care visit

### 

```{r}
# Install haven if you don't have it already
install.packages("haven")

# Load the haven package
library(haven)

# Read a .sas7bdat file
data <- read_sas("/Users/molabisi/Downloads/PRAMS ARF/phase8_arf_2016_2022.sas7bdat")

# View the first few rows of the data
head(data)

```

```{r}

#clean data
## count/how many people in data set
cat("Total rows in the dataset:", nrow(data), "\n")

##load libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(caret)
library(boot)
install.packages("tableone")
library(tableone)

## make exposure variables more concise
data_clean <- data %>%
  mutate(
    # For PRE_VIT_RAW (Pre-preg -- folic acid)
    PRE_VIT_RAW = case_when(
      PRE_VIT_RAW == 1 ~ 0,   # NO -> 0
      PRE_VIT_RAW == 2 ~ 1,   # YES -> 1
      TRUE ~ PRE_VIT_RAW # Keep other values unchanged
    ),
    
    # For VPP_VIT_RAW (PPV -- folic acid)
    VPP_VIT_RAW = case_when(
      VPP_VIT_RAW == 1 ~ 0,   # NO -> 0
      VPP_VIT_RAW == 2 ~ 1,   # YES -> 1
      TRUE ~ VPP_VIT_RAW # Keep other values unchanged
    ),
    
    # Convert to factors after recoding
    PRE_VIT_RAW = factor(PRE_VIT_RAW, levels = c(0, 1)),
    VPP_VIT_RAW = factor(VPP_VIT_RAW, levels = c(0, 1))
  )

library(dplyr)
## make outcome variables more concise
data_clean <- data_clean %>%
  mutate(
    # Simplify birth weight categories into broader groups
    GRAM_NAPHSIS = case_when(
      GRAM_NAPHSIS %in% c(125, 375, 625, 875, 1125, 1375) ~ 0,   # Less than 1500 grams VLBW
      GRAM_NAPHSIS %in% c(1625, 1875, 2125, 2375) ~ 1,  # Less than 2500 grams LBW
      GRAM_NAPHSIS %in% c(2625, 2875, 3125, 3375, 3625, 3875) ~ 2, # Between 2500 and 4000 grams (NBW)
      GRAM_NAPHSIS >= 4125 ~ 3, # Over 4000 grams (MACROSOMIA)
      TRUE ~ NA_real_  # Missing or unknown values will remain NA
    ),
    GRAM_NAPHSIS = factor(GRAM_NAPHSIS, levels = c(0, 1, 2, 3))
  )

##make covariates more concise
data_clean <- data_clean %>%
  mutate(
    # Maternal Age
    MAT_AGE_PU = case_when(
      MAT_AGE_PU == 17 ~ 0,      # Age 17 or younger
      MAT_AGE_PU == 45 ~ 1,      # Age 45 or older
      TRUE ~ NA_real_          # Other ages remain unchanged
    ),
    
    # Parity
    PRE_LB_NAPHSIS = case_when(
      PRE_LB_NAPHSIS == 0 ~ 0,       # No children (0)
      PRE_LB_NAPHSIS == 1 ~ 1,       # One child (1)
      PRE_LB_NAPHSIS %in% c(2, 3, 4) ~ 2, # 2+ children (Yes, multiple)
      TRUE ~ NA_real_
    ),
    
    # BMI (e.g., categorized into underweight, normal, overweight, obese)
    MOM_BMIG_QX_REV = case_when(
      MOM_BMIG_QX_REV == 1 ~ 0, #"Underweight (<18.5)",
      MOM_BMIG_QX_REV == 2 ~ 1, #"Normal weight (18.5-24.9)",
      MOM_BMIG_QX_REV == 3 ~ 2, #"Overweight (25.0-29.9)",
      MOM_BMIG_QX_REV == 4 ~ 3, #"Obese (30+)",
      TRUE ~ NA_real_
    ),
    
   # Education: Convert to simpler categories
    MAT_ED = case_when(
      MAT_ED == 1 ~ 0,  # <= 8th grade
      MAT_ED == 2 ~ 1,  # 9-12th grade, no diploma
      MAT_ED == 3 ~ 2,  # High school graduate / GED
      MAT_ED == 4 ~ 3,  # Some college, no degree
      MAT_ED == 5 ~ 4,  # Bachelor's, Master's, or higher
      TRUE ~ NA_real_
    ),

    # Income: Group income into broader categories
    INCOME8 = case_when(
      INCOME8 %in% c(1, 2, 3, 4) ~ 0,  # Low-income group (<=$32,000)
      INCOME8 %in% c(5, 6, 7, 8) ~ 1,  # Middle-income group ($32,001-$57,000)
      INCOME8 %in% c(9, 10, 11, 12) ~ 2,  # Upper-middle-income group ($57,001-$85,000)
      INCOME8 >= 13 ~ 3,  # High-income group (> $85,001)
      TRUE ~ NA_real_
    )

  ) %>%
  mutate(
    # Convert categorical variables to factors after recoding
    MAT_AGE_PU = factor(MAT_AGE_PU, levels = c(0, 1)),
    PRE_LB_NAPHSIS = factor(PRE_LB_NAPHSIS, levels = c(0, 1, 2)),
    MOM_BMIG_QX_REV = factor(MOM_BMIG_QX_REV, levels = c(0, 1, 2, 3)),
    MAT_ED = factor(MAT_ED, levels = c(0, 1, 2, 3, 4)),
    INCOME8 = factor(INCOME8, levels = c(0, 1, 2, 3))
  )

library(dplyr)

# Creating a new category for regions
data_clean <- data_clean %>%
  mutate(STATE = case_when(
    STATE %in% c("NY", "NJ", "MA", "PA", "RI", "CT", "VT", "NH", "ME", "YC") ~ 0, #"Northeast",
    STATE %in% c("IL", "IN", "OH", "MI", "WI", "MN", "IA", "MO", "KS", "NE", "SD", "ND") ~ 1, # "Midwest",
    STATE %in% c("DE", "MD", "VA", "KY", "TN", "NC", "SC", "GA", "FL", "AL", "MS", "LA", "AR", "OK", "TX", "WV") ~ 2, #"South",
    STATE %in% c("WA", "OR", "CA", "CO", "AZ", "NM", "UT", "NV", "MT", "WY", "ID", "AK", "HI") ~ 3, #"West",
    STATE == c("DC", "PR") ~ 4, #"Other",  # District of Columbia # Puerto Rico
    TRUE ~ NA_real_ 
  )) %>%
  mutate(STATE = factor(STATE, levels = c(0, 1, 2, 3, 4)))

data_clean <- data_clean %>%
  mutate(PNC_VST = case_when(
    PNC_VST == 0 ~ 0, # No visits
    PNC_VST >= 1 & PNC_VST <= 5 ~ 1, # 1-5 visits
    PNC_VST >= 6 & PNC_VST <= 10 ~ 2, # 6-10 visits
    PNC_VST >= 11 & PNC_VST <= 15 ~ 3, # 11-15 visits
    PNC_VST >= 16 & PNC_VST <= 20 ~ 4, # 16-20 visits
    PNC_VST >= 21 & PNC_VST <= 30 ~ 5, # 21-30 visits
    PNC_VST >= 31 & PNC_VST <= 40 ~ 6, # 31-40 visits
    PNC_VST > 40 ~ 7, # More than 40 visits
    TRUE ~ NA_real_ # Catch any unexpected values (use NA_real_ for numeric)
  )) %>%
  mutate(PNC_VST = factor(PNC_VST, 
                          levels = c(0, 1, 2, 3, 4, 5, 6, 7)))


table(data_clean$PRE_VIT_RAW)
table(data_clean$VPP_VIT_RAW)
table(data_clean$GRAM_NAPHSIS)
table(data_clean$MAT_AGE_PU)
table(data_clean$PRE_LB_NAPHSIS)
table(data_clean$MOM_BMIG_QX_REV)
table(data_clean$MAT_ED)
table(data_clean$STATE)
table(data_clean$INCOME8)
table(data_clean$PNC_VST)
```

```{r}
# variables I want in the descriptive table
variables <- c("PRE_VIT_RAW", "VPP_VIT_RAW", "GRAM_NAPHSIS", "MAT_AGE_PU", "PRE_LB_NAPHSIS", 
               "MOM_BMIG_QX_REV", "MAT_ED", "STATE", "INCOME8", "PNC_VST")

# Create the descriptive table
table <- CreateTableOne(vars = variables, data = data_clean, factorVars = c("PRE_VIT_RAW", "VPP_VIT_RAW", "GRAM_NAPHSIS", 
                                                                           "MAT_AGE_PU", "PRE_LB_NAPHSIS", "MOM_BMIG_QX_REV", 
                                                                           "MAT_ED", "STATE", "INCOME8", "PNC_VST"))

# Print the table
print(table, showAllLevels = TRUE)
```

```{r}
# Variables to include in the descriptive table
variables <- c("PRE_VIT_RAW", "VPP_VIT_RAW", "GRAM_NAPHSIS", "MAT_AGE_PU", "PRE_LB_NAPHSIS", 
               "MOM_BMIG_QX_REV", "MAT_ED", "STATE", "INCOME8", "PNC_VST")

# descriptive table by the outcome variable GRAM_NAPHSIS
table <- CreateTableOne(vars = variables, 
                        data = data_clean, 
                        factorVars = c("PRE_VIT_RAW", "VPP_VIT_RAW", "GRAM_NAPHSIS", 
                                       "MAT_AGE_PU", "PRE_LB_NAPHSIS", "MOM_BMIG_QX_REV", 
                                       "MAT_ED", "STATE", "INCOME8", "PNC_VST"),
                        strata = "GRAM_NAPHSIS")  # Stratifying by GRAM_NAPHSIS

# Print the table
print(table, showAllLevels = TRUE)
```

```{r}
model_logistic2 <- glm(GRAM_NAPHSIS ~ PRE_VIT_RAW, 
                           data = data_clean, 
                           family = binomial)
summary(model_logistic2)

model_logistic3 <- glm(GRAM_NAPHSIS ~ VPP_VIT_RAW, 
                           data = data_clean, 
                           family = binomial)
summary(model_logistic3)

##have to figure out why im getting this warning 
##Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
model_logistic <- glm(GRAM_NAPHSIS ~ PRE_VIT_RAW + MAT_AGE_PU + PRE_LB_NAPHSIS + 
                           MOM_BMIG_QX_REV + MAT_ED + STATE + INCOME8 + PNC_VST, 
                           data = data_clean, 
                           family = binomial)
summary(model_logistic)

model_logistic1 <- glm(GRAM_NAPHSIS ~ VPP_VIT_RAW + MAT_AGE_PU + PRE_LB_NAPHSIS + 
                           MOM_BMIG_QX_REV + MAT_ED + STATE + INCOME8 + PNC_VST, 
                           data = data_clean, 
                           family = binomial)
summary(model_logistic1)
```

######################
```{r}
library(nnet)

# For the first multinomial model (model_multinom)
model_multinom <- multinom(GRAM_NAPHSIS ~ PRE_VIT_RAW + + MAT_AGE_PU + PRE_LB_NAPHSIS + 
                           MOM_BMIG_QX_REV + MAT_ED + STATE + INCOME8 + PNC_VST, 
                           data = data_clean)

# Exponentiate the coefficients to get the odds ratios (ORs)
odds_ratios_multinom <- exp(coef(model_multinom))

# Calculate 95% confidence intervals for the ORs
confint_multinom <- exp(confint(model_multinom))

# Print the odds ratios and confidence intervals
odds_ratios_multinom
confint_multinom

# For the second multinomial model (model_multinom1)
model_multinom1 <- multinom(GRAM_NAPHSIS ~ VPP_VIT_RAW + MAT_AGE_PU + PRE_LB_NAPHSIS + 
                           MOM_BMIG_QX_REV + MAT_ED + STATE + INCOME8 + PNC_VST, 
                            data = data_clean)

# Exponentiate the coefficients to get the odds ratios (ORs)
odds_ratios_multinom1 <- exp(coef(model_multinom1))

# Calculate 95% confidence intervals for the ORs
confint_multinom1 <- exp(confint(model_multinom1))

# Print the odds ratios and confidence intervals
odds_ratios_multinom1
confint_multinom1

```



```{r}

model_interaction <- glm(GRAM_NAPHSIS ~ PRE_VIT_RAW:PNC_VST + + MAT_AGE_PU + PRE_LB_NAPHSIS + 
                           MOM_BMIG_QX_REV + MAT_ED + STATE + INCOME8 + PNC_VST, 
                           data = data_clean, family = binomial())

# Summarize the model results
summary(model_interaction)
```

## Results {#sec-results}

Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

## Conclusion

This the conclusion. The @sec-results can be invoked here.

1.  Organization WH. International statistical classification of diseases and related health problems, tenth revision, 2nd ed. World Health Organization; 2004.

2.  Jańczewska, I., Wierzba, J., Jańczewska, A., Szczurek-Gierczak, M., & Domżalska-Popadiuk, I. (2023). Prematurity and Low Birth Weight and Their Impact on Childhood Growth Patterns and the Risk of Long-Term Cardiovascular Sequelae. *Children*, *10*(10), Article 10. <https://doi.org/10.3390/children10101599>

3.  Fu, M., Song, W., Yu, G., Yu, Y., & Yang, Q. (2023). Risk factors for length of NICU stay of newborns: A systematic review. *Frontiers in Pediatrics*, *11*, 1121406. <https://doi.org/10.3389/fped.2023.1121406>

4.  Houk, V. N., Oakley, G. P., Erickson, J. D., Mulinare, J., & James, L. M. (n.d.). *Recommendations for the use of folic acid to reduce the number of cases of spina bifida and other neural tube defects*. Retrieved November 12, 2024, from <https://stacks.cdc.gov/view/cdc/7079>

5.  Scholl, T. O., & Johnson, W. G. (2000). Folic acid: Influence on the outcome of pregnancy1234. *The American Journal of Clinical Nutrition*, *71*(5), 1295S-1303S. <https://doi.org/10.1093/ajcn/71.5.1295s>
